---
title: "METABRIC"
params:
  data: "../data"
---

Analysis of immunological signatures in METABRIC
- Part 1: Prepare data
- Part 2: Analysis

Running the code in part 1 will download data associated with METABRIC and save them to *params$data/mb.rds*.
Running the code in part 2 will save the analysis results to *params$data/mb_results.rds*.
To use CIBERSORT, an academic license needs to be obtained and the script and signature files should be placed into *lib/cibersort*.

```{r setup}
library(ggplot2)
library(patchwork)
library(dplyr)
library(immunedeconv)
library(survival)
library(limma)
library(matrixStats)
bcra_env <- new.env()
source(file.path("..", "lib", "r-functions.R"), local = bcra_env) |> invisible()
attach(bcra_env)
dir.create(params$data, showWarnings = FALSE)
set_cibersort_binary(file.path("..", "lib", "cibersort", "CIBERSORT.R"))
set_cibersort_mat(file.path("..", "lib", "cibersort", "LM22.txt"))
knitr::opts_chunk$set(fig.width = 10, fig.height = 4)
```


#### Prepare Data

Clinical and expression data
```{r}
# cbioportal
if (!dir.exists(fp(params$data, "METABRIC", "brca_metabric"))){
  download.file(url = "https://cbioportal-datahub.s3.amazonaws.com/brca_metabric.tar.gz",
                destfile = fp(params$data, "METABRIC", "brca_metabric.tar.gz"), method = "curl")
  untar(tarfile = fp(params$data, "METABRIC", "brca_metabric.tar.gz"), exdir = fp(params$data, "METABRIC"))
  unlink(fp(params$data, "METABRIC", "brca_metabric.tar.gz"))
}

# clincal follow-up data from Rueda et al, 2019
if (!file.exists(fp(params$data, "METABRIC", "rueda_clinical", "Rueda_ST5.txt"))){
  dir.create(fp(params$data, "METABRIC", "rueda_clinical"), showWarnings = FALSE)
  download.file(url = "https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-019-1007-8/MediaObjects/41586_2019_1007_MOESM7_ESM.txt",
              destfile = fp(params$data, "METABRIC", "rueda_clinical", "Rueda_ST5.txt"))
  download.file("https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-019-1007-8/MediaObjects/41586_2019_1007_MOESM10_ESM.xls",
              destfile = fp(params$data, "METABRIC", "rueda_clinical", "Rueda_ST5_info.xls"))
}

mb <- load_metabric_data(fp(params$data, "METABRIC"))
mb$metadata <- load_metabric_metadata(fp(params$data, "METABRIC"))
colnames(mb$metadata) <- colnames(mb$metadata) |> tolower()
mb$metadata$ID <- rownames(mb$metadata)
mb$metadata <- mb$metadata |> dplyr::rename(subcohort = cohort, er_status_by_ihc = er_ihc, pr_status_by_ihc = pr_status, her2_status_by_ihc = her2_status)
mb$expr <- merge_rows(dplyr::select(mb$expr, -c(Hugo_Symbol, Entrez_Gene_Id)), ids = mb$expr$Hugo_Symbol, FUN = \(x) mean(x, na.rm = TRUE))
mb$expr_zscores <- merge_rows(dplyr::select(mb$expr_zscores, -c(Hugo_Symbol, Entrez_Gene_Id)), ids = mb$expr_zscores$Hugo_Symbol, FUN = \(x) mean(x, na.rm = TRUE))
```

Mutations and TMB
```{r}
mb_maf <- mb$mutations |> maftools::read.maf(isTCGA = FALSE)
mb_tmb <- maftools::tmb(mb_maf, captureSize = 1)
mb$tmb <- mb_tmb |> dplyr::rename(ID = Tumor_Sample_Barcode, TMB_rate_maf = total_perMB, TMB_count_maf = total)
mb$metadata$TMB_rate_maf <- NULL
mb$metadata$TMB_count_maf <- NULL
mb$metadata <- dplyr::left_join(x = mb$metadata, y = dplyr::select(mb$tmb, TMB_rate_maf, TMB_count_maf, ID), by = "ID")
rownames(mb$metadata) <- mb$metadata$ID
mb$metadata$TMB3 <- ifelse(mb$metadata$TMB_rate_maf > 3, ">3mutpMB", "<3mutpMB")
mb$metadata$TMB10 <- ifelse(mb$metadata$TMB_count_maf > 10, ">10mutpMB", "<10mutpMB")

# use the available TMB:
mb$metadata$TMB_count <- mb$metadata$tmb_nonsynonymous 
mb$metadata$TMB_count_median <- stratify(mb$metadata$TMB_count)
mb$metadata$TMB_count_group <- ifelse(mb$metadata$TMB_count > quantile(mb$metadata$TMB_count, probs = 0.9, na.rm = TRUE), "high (>90%)", "low (<90%)")
```

Survival definitions
```{r}
# Overall survival
mb$os <- data.frame(row.names = rownames(mb$metadata))
mb$os$time <- mb$metadata$os_months
mb$os$status <- NA
mb$os$status[mb$metadata$vital_status %in% c("Died of Disease", "Died of Other Causes")] <- 1
mb$os$status[mb$metadata$vital_status %in% "Living"] <- 0
mb$metadata$os_status

# Disease-specific survival
mb$dss <- data.frame(row.names = rownames(mb$metadata))
mb$dss$time <- mb$metadata$os_months
mb$dss$status <- NA
mb$dss$status[mb$metadata$vital_status %in% "Died of Other Causes"] <- 0
mb$dss$status[mb$metadata$vital_status %in% "Died of Disease"] <- 1
mb$dss$status[mb$metadata$vital_status %in% "Living"] <- 0

# Recurrence-free survival
mb$dfs <- data.frame(row.names = rownames(mb$metadata))
mb$dfs$status = sub(":.*", "",  mb$metadata$rfs_status) |> as.numeric()
mb$dfs$time = mb$metadata$rfs_months

# we can check in more detail how to get these data:
mb$dfs_times <- data.frame(row.names = rownames(mb$metadata))
mb$dfs_events <- data.frame(row.names = rownames(mb$metadata))
mb$dfs_events$dss_death <- mb$dss$status
mb$dfs_events$dss_death[is.na(mb$dfs_events$dss_death)] <- 0
mb$dfs_events$lr <- mb$metadata$lr
mb$dfs_events$dr <- mb$metadata$dr
mb$dfs_status <- as.numeric(rowSums(mb$dfs_events) > 0)
mb$dfs_times$dss_time <- mb$dss$time
mb$dfs_times$tlr <- mb$metadata$tlr / 30 # originally 30 was, so here we use the same
mb$dfs_times$tdr <- mb$metadata$tdr / 30
mb$dfs_time <- matrixStats::rowMins(data.matrix(mb$dfs_times), na.rm = TRUE)
mb$dfs_time[is.infinite(mb$dfs_time)] <- NA

# convert to years
mb$os_years <- mb$os |> dplyr::mutate(time = time * 30 / 365.2425)
mb$dss_years <- mb$dss |> dplyr::mutate(time = time * 30 / 365.2425)
mb$dfs_years <- mb$dfs |> dplyr::mutate(time = time * 30 / 365.2425)
```

Receptor subtype definitions
```{r}
mb$metadata$er_status_by_ihc[mb$metadata$er_status_by_ihc == "Positve"] <- "Positive" # typo

# TNBC definition from https://doi.org/10.1038%2Fs41598-021-83913-7
mb$metadata$er_status_by_ihc[mb$metadata$er_status_by_ihc == ""] <- NA
tnbc_cases <- subset(mb$metadata, er_status_by_ihc == "Negative" & her2_status_by_ihc == "Negative" & her2_snp6 != "GAIN" & pr_status_by_ihc == "Negative")
tnbc_cases <- subset(tnbc_cases, histological_subtype == "Ductal/NST") # none
tnbc_cases <- subset(tnbc_cases, er_status == "Negative") 
mb$metadata$TNBC <- rownames(mb$metadata) %in% rownames(tnbc_cases)
sum(mb$metadata$TNBC) # this would result in 199 cases, as in paper
mb$metadata$ER_status <- ifelse(mb$metadata$er_status_by_ihc %in% "Positive", "ER+", "ER?")
mb$metadata$ER_status[mb$metadata$er_status_by_ihc == "Negative"] <- "ER-"
mb$metadata$PR_status <- ifelse(mb$metadata$pr_status_by_ihc %in% "Positive", "PR+", "PR?")
mb$metadata$PR_status[mb$metadata$pr_status_by_ihc == "Negative"] <- "PR-"
mb$metadata$HER2_status <- "HER2?"
mb$metadata$HER2_status[mb$metadata$her2_status_by_ihc == "Positive" & mb$metadata$her2_snp6 != "LOSS"] <- "HER2+"
mb$metadata$HER2_status[mb$metadata$her2_status_by_ihc == "Negative" & mb$metadata$her2_snp6 != "GAIN"] <- "HER2-"

# combine
mb$metadata <- mb$metadata |> mutate(receptor_status = paste0(ER_status, PR_status, HER2_status))
mb$metadata$receptor_status |> table() |> sort(decreasing = TRUE)

# IHC three-group (HR+HER2-; HER2+; TNBC)
mb$metadata$receptor_subtype <- "other"
mb$metadata$receptor_subtype[mb$metadata$HER2_status == "HER2+"] <- "HER2+"
mb$metadata$receptor_subtype[mb$metadata$HER2_status == "HER2-" & (mb$metadata$ER_status == "ER+" | mb$metadata$PR_status == "PR+")] <- "HR+HER2-"
#mb$metadata$receptor_subtype[mb$metadata$HER2_status == "HER2-" & mb$metadata$ER_status == "ER-"] <- "TNBC"
mb$metadata$receptor_subtype[mb$metadata$receptor_status == "ER-PR-HER2-"] <- "TNBC"
mb$metadata$TNBC <- ifelse(mb$metadata$receptor_subtype == "TNBC", "yes", "no")
```

Discovery/validation cohort and PAM50_subtype
```{r}
if (!file.exists(fp(params$data, "METABRIC", "other", "Curtis_ST"))){
  download.file(url = "https://static-content.springer.com/esm/art%3A10.1038%2Fnature10983/MediaObjects/41586_2012_BFnature10983_MOESM266_ESM.zip",
                destfile = fp(params$data, "METABRIC", "other", "41586_2012_BFnature10983_MOESM266_ESM.zip"))
    system(paste0("unzip ", fp(params$data, "METABRIC", "other", "41586_2012_BFnature10983_MOESM266_ESM.zip"), " -d ",
                fp(params$data, "METABRIC", "other", "Curtis_ST")), intern = TRUE, ignore.stdout = TRUE)
}

mb_discovery = read.delim(fp(params$data, "METABRIC", "other", "Curtis_ST", "2011-04-05112E-Tables_S1-S6", "table_S2_revised.txt"))
mb_discovery$METABRIC_ID <- sub("-", "", mb_discovery$METABRIC_ID)
mb_validation = read.delim(fp(params$data, "METABRIC", "other", "Curtis_ST", "2011-04-05112E-Tables_S1-S6", "table_S3_revised.txt"))
mb_validation$METABRIC_ID <- sub("-", "", mb_validation$METABRIC_ID)
mb$metadata$cohort <- "undef"
mb$metadata$cohort[mb$metadata$ID %in% mb_discovery$METABRIC_ID] <- "discovery"
mb$metadata$cohort[mb$metadata$ID %in% mb_validation$METABRIC_ID] <- "validation"

mb$metadata$PAM50_subtype <- NA
mb$metadata[mb_discovery$METABRIC_ID,]$PAM50_subtype <- mb_discovery$Pam50Subtype
mb$metadata[mb_validation$METABRIC_ID,]$PAM50_subtype <- mb_validation$Pam50Subtype
mb$metadata$PAM50_subtype[mb$metadata$PAM50_subtype == "Her2"] <- "Her2e"
mb$metadata$PAM50_subtype[mb$metadata$PAM50_subtype == "NC"] <- NA
```

Recode
```{r}
mb$metadata[mb$metadata == ""] <- NA
mb$metadata$cellularity <- mb$metadata$cellularity |> tolower()
mb$metadata$chemotherapy <- mb$metadata$chemotherapy |> tolower()
mb$metadata$er_status_by_ihc <- mb$metadata$er_status_by_ihc |> tolower()
mb$metadata$her2_snp6 <- mb$metadata$her2_snp6 |> tolower()
mb$metadata$hormone_therapy <- mb$metadata$hormone_therapy |> tolower()
mb$metadata$inferred_menopausal_state <- mb$metadata$inferred_menopausal_state |> tolower()
mb$metadata$sex <- mb$metadata$sex |> tolower()
mb$metadata$laterality <- mb$metadata$laterality |> tolower()
mb$metadata$radio_therapy <- mb$metadata$radio_therapy |> tolower()
mb$metadata$breast_surgery <- mb$metadata$breast_surgery |> tolower()
mb$metadata$cancer_type_detailed <- mb$metadata$cancer_type_detailed |> tolower()
mb$metadata$er_status <- mb$metadata$er_status |> tolower()
mb$metadata$er_status <- mb$metadata$er_status |> tolower()
mb$metadata$her2_status_by_ihc <- mb$metadata$her2_status_by_ihc |> tolower()
mb$metadata$pr_status_by_ihc <- mb$metadata$pr_status_by_ihc |> tolower()
mb$metadata$sample_type <- mb$metadata$sample_type |> tolower()
mb$metadata$pr_status_by_ihc <- mb$metadata$pr_status_by_ihc |> tolower()
mb$metadata$ct <- mb$metadata$ct |> tolower()
mb$metadata$ht <- mb$metadata$ht |> tolower()
mb$metadata$rt <- mb$metadata$rt |> tolower()
mb$metadata$complete_rec_history <- mb$metadata$complete_rec_history |> tolower()

# claudin subtypes
mb$metadata$claudin_subtype[mb$metadata$claudin_subtype == "NC"] <- NA # not classified
mb$metadata$claudin_subtype[mb$metadata$claudin_subtype == "claudin-low"] <- "Claudin-low"
mb$metadata$claudin_subtype[mb$metadata$claudin_subtype == "Her2"] <- "Her2e"
mb$metadata$claudin_subtype |>  table(useNA = "ifany") |> sort(decreasing = TRUE)

# therapy (missing for 536 patients)
mb$metadata$chemotherapy |>  table(useNA = "ifany")
mb$metadata$hormone_therapy |>  table(useNA = "ifany")
mb$metadata$radio_therapy |>  table(useNA = "ifany")
mb$metadata$breast_surgery |>  table(useNA = "ifany")

mb$metadata$therapy <- paste(ifelse(mb$metadata$chemotherapy == "yes", "CT", ""),
                             ifelse(mb$metadata$hormone_therapy == "yes", "HT", ""),
                             ifelse(mb$metadata$radio_therapy == "yes", "RT", ""), sep = "+")
mb$metadata$therapy <- mb$metadata$therapy |>
  gsub(pattern = "NA", replacement = "") |>
  gsub(pattern = "^\\+*", replacement = "") |>
  gsub(pattern = "++", replacement = "+", fixed = TRUE) |>
  gsub(pattern = "\\+$", replacement = "")
mb$metadata$therapy[is.na(mb$metadata$chemotherapy) | is.na(mb$metadata$hormone_therapy) | is.na(mb$metadata$radio_therapy)] <- NA
mb$metadata$therapy[mb$metadata$therapy == ""] <- "none"

# histological subtypes
mb$metadata$histological_type |>  table(useNA = "ifany") |> sort(decreasing = TRUE)
mb$metadata$histological_subtype_full <- mb$metadata$histological_subtype
mb$metadata$histological_subtype <- NA
mb$metadata$histological_subtype[mb$metadata$histological_subtype_full == "Medullary"] <- "MC"
mb$metadata$histological_subtype[mb$metadata$histological_subtype_full == "Mucinous"] <- "MUC"
mb$metadata$histological_subtype[mb$metadata$histological_subtype_full == "Metaplastic"] <- "MPC"
mb$metadata$histological_subtype[mb$metadata$histological_subtype_full == "Lobular"] <- "ILC"
mb$metadata$histological_subtype[mb$metadata$histological_subtype_full == "Ductal/NST"] <- "IDC"
mb$metadata$histological_subtype[mb$metadata$histological_subtype_full == "Tubular/ cribriform"] <- "IDC-TC"
mb$metadata$histological_subtype[mb$metadata$histological_subtype_full == "Mixed"] <- "other"
mb$metadata$histological_subtype[mb$metadata$histological_subtype_full == "Other"] <- "other"
mb$metadata$histological_subtype |>  table(useNA = "ifany") |> sort(decreasing = TRUE)
mb$metadata$stage <- mb$metadata$tumor_stage
mb$metadata <- mb$metadata |> dplyr::rename(NPI = npi, normal_sample_id = matched_normal_metabric_id, age = age_at_diagnosis, menopause_status = inferred_menopausal_state)
```

Deconvolution
```{r}
# prepare expression data
impute_values <- rep(rowMins(data.matrix(mb$expr), na.rm = TRUE), ncol(mb$expr)) |> matrix(ncol =  ncol(mb$expr))
mb$expr[is.na(mb$expr)] <- impute_values[is.na(mb$expr)]
mb$pseudotpm <- 2^mb$expr

# cibersort
mb$cibersort_rel <- deconvolute_cibersort(mb$pseudotpm, perm=1000, QN = TRUE, absolute = FALSE, arrays = TRUE) 
mb$cibersort_rel <- data.frame(t(mb$cibersort_rel))
colnames(mb$cibersort_rel) <- gsub(".", "_", colnames(mb$cibersort_rel), fixed = TRUE)
colnames(mb$cibersort_rel) <- gsub("__Tregs", "", colnames(mb$cibersort_rel))
colnames(mb$cibersort_rel) <- gsub("_$", "", colnames(mb$cibersort_rel))
mb$cibersort_rel$Correlation <- NULL
mb$cibersort_rel$RMSE <- NULL

# estimate
mb$estimate <- immunedeconv::deconvolute_estimate(mb$pseudotpm) |> as.data.frame()
mb$estimate <- data.frame(t(mb$estimate))
```

```{r}
MHC <- get_mhc_genes()
mb$metadata$HLA_I_mRNAs <- colSums(mb$expr[MHC$classI,], na.rm = TRUE)[mb$metadata$ID]
mb$metadata$HLA_II_mRNAs <- colSums(mb$expr[MHC$classII,], na.rm = TRUE)[mb$metadata$ID]
mb$metadata$B2M_mRNA <- colSums(mb$expr["B2M",], na.rm = TRUE)[mb$metadata$ID]
mb$metadata$PDL1_mRNA <- colSums(mb$expr["CD274",], na.rm = TRUE)[mb$metadata$ID]
mb$metadata$CD4_mRNA <- colSums(mb$expr["CD4",], na.rm = TRUE)[mb$metadata$ID]
mb$metadata$CD8A_mRNA <- colSums(mb$expr["CD8A",], na.rm = TRUE)[mb$metadata$ID]
mb$metadata$GZMA_mRNA <- colSums(mb$expr["GZMA",], na.rm = TRUE)[mb$metadata$ID]

hist(mb$metadata$B2M_mRNA, 50)
mb$metadata$B2M_mRNA_group <- ifelse(mb$metadata$B2M_mRNA > quantile(mb$metadata$B2M_mRNA, probs = 0.1, na.rm = TRUE), "normal (>10%)", "low (<10%)")
hist(mb$metadata$HLA_I_mRNAs, 50)
mb$metadata$HLA_I_mRNAs_group <- ifelse(mb$metadata$HLA_I_mRNAs > quantile(mb$metadata$HLA_I_mRNAs, probs = 0.1, na.rm = TRUE), "normal (>10%)", "low (<10%)")
hist(mb$metadata$HLA_II_mRNAs, 50)
mb$metadata$HLA_II_mRNAs_group <- ifelse(mb$metadata$HLA_II_mRNAs > quantile(mb$metadata$HLA_II_mRNAs, probs = 0.1, na.rm = TRUE), "normal (>10%)", "low (<10%)")
hist(mb$metadata$PDL1_mRNA, 50)
mb$metadata$PDL1_mRNA_group <- ifelse(mb$metadata$PDL1_mRNA > quantile(mb$metadata$PDL1_mRNA, probs = 0.9, na.rm = TRUE), "high (>90%)", "normal (<90%)")
mb$metadata <- mb$metadata |> dplyr::select(-c(metabric_id, sample_id, cancer_type, sample_type, tumor_stage, histological_subtype_full))
```

Data cleanup
```{r}
# combine deconvolution results
decon_res <- Reduce(f = function(x,y){ dplyr::full_join(x,y,by="ID")}, x = list(
  mb$estimate |> dplyr::select(StromalScore, ImmuneScore) %>% mutate(ID = rownames(.)),
  mb$cibersort_rel |> dplyr::rename_with( ~ paste0("Cbs_", .x)) %>% mutate(ID = rownames(.))
))
rownames(decon_res) <- decon_res$ID
decon_res <- decon_res |> dplyr::rename(Cbs_B_memory = Cbs_B_cells_memory,
                                        Cbs_CD8_T_cells = Cbs_T_cells_CD8,
                                        Cbs_CD4_T_naive = Cbs_T_cells_CD4_naive,
                                        Cbs_CD4_T_mem_resting = Cbs_T_cells_CD4_memory_resting,
                                        Cbs_CD4_T_mem_activated = Cbs_T_cells_CD4_memory_activated,
                                        Cbs_T_fh = Cbs_T_cells_follicular_helper,
                                        Cbs_T_reg = Cbs_T_cells_regulatory,
                                        Cbs_T_gd = Cbs_T_cells_gamma_delta,
                                        Cbs_NK_resting = Cbs_NK_cells_resting,
                                        Cbs_NK_activated = Cbs_NK_cells_activated,
                                        Cbs_DC_resting = Cbs_Dendritic_cells_resting,
                                        Cbs_DC_activated = Cbs_Dendritic_cells_activated)
mb$vars$celltypes <- colnames(decon_res)

df <- dplyr::inner_join(mb$metadata, decon_res, by="ID")
df <- df |> dplyr::select(-starts_with(c("os", "dss", "dfs")), -patient_id, -ends_with("_barcode"))
df <- df[,df |> sapply(function(x) sum(!is.na(x))>1 )]

tab_receptor_status <- df$receptor_status |>  table(useNA = "ifany") |> sort(decreasing = TRUE)
df$receptor_status[grepl("?", df$receptor_status, fixed = TRUE)] <- NA
df$receptor_status[tab_receptor_status[df$receptor_status] < 40] <- NA
df$receptor_status |>  table(useNA = "ifany") |> sort(decreasing = TRUE)
df$receptor_status[df$receptor_status == "ER+PR+HER2-"] <- "E+P+"
df$receptor_status[df$receptor_status == "ER+PR-HER2-"] <- "E+"
df$receptor_status[df$receptor_status == "ER-PR-HER2-"] <- "---"
df$receptor_status[df$receptor_status == "ER-PR-HER2+"] <- "H+"
df$receptor_status[df$receptor_status == "ER+PR-HER2+"] <- "E+H+"
df$receptor_status[df$receptor_status == "ER+PR+HER2+"] <- "E+P+H+"
df$receptor_status |>  table(useNA = "ifany") |> sort(decreasing = TRUE)

df$log_TMB_rate <- log10(df$TMB_rate + 1)
df$TMB_rate <- NULL

df$os <- Surv(mb$os[df$ID,]$time, mb$os[df$ID,]$status)
df$dss <- Surv(mb$dss[df$ID,]$time, mb$dss[df$ID,]$status)
df$dfs <- Surv(mb$dfs[df$ID,]$time, mb$dfs[df$ID,]$status)

vars_celltypes <- setdiff(colnames(decon_res), c("ID", "StromalScore", "ImmuneScore", "Cbs_P_value"))
ix <- vars_celltypes |> sub(pattern = "Bp_|Cbs_|Qs_", replacement = "") |> order()
vars_celltypes <- c("ImmuneScore", "StromalScore", grep("HE_", colnames(df), value = TRUE), vars_celltypes[ix])
names(vars_celltypes) <- vars_celltypes

df$ER_status[grep("?", df$ER_status, fixed = TRUE)] <- NA
df$ER_status <- factor(df$ER_status, ordered = TRUE, levels = c("ER-", "ER+")) # if ordered
df$ER_status <- relevel(factor(df$ER_status, ordered = FALSE), ref = "ER-") # if not ordered

df$PR_status[grep("?", df$PR_status, fixed = TRUE)] <- NA
df$PR_status <- factor(df$PR_status, ordered = TRUE, levels = c("PR-", "PR+"))
df$PR_status <- relevel(factor(df$PR_status, ordered = FALSE), ref = "PR-")

df$HER2_status[grep("?", df$HER2_status, fixed = TRUE)] <- NA
df$HER2_status <- factor(df$HER2_status, ordered = TRUE, levels = c("HER2-", "HER2+"))
df$HER2_status <- relevel(factor(df$HER2_status, ordered = FALSE), ref = "HER2-")

df$menopause_status <- factor(df$menopause_status, ordered = TRUE, levels = c("pre", "post"))
df$stage_hi <- factor(ifelse(df$stage < 3, "low", "high"), ordered = TRUE, levels = c("low", "high"))
df$stage_hi[df$stage == 0] <- NA
df$stage_hi <- relevel(factor(df$stage_hi, ordered = FALSE), ref = "low")

df$cellularity_hi <- factor(ifelse(df$cellularity %in% c("low", "moderate"), "low", "high"), ordered = TRUE, levels = c("low", "high"))
df$cellularity_hi[is.na(df$cellularity)] <- NA
df$cellularity_hi <- relevel(factor(df$cellularity_hi, ordered = FALSE), ref = "low")

df$laterality <- factor(ifelse(df$breast_tumour_laterality == "l", "left", "right"), ordered = TRUE, levels = c("left", "right"))
df$laterality[is.na(df$breast_tumour_laterality)] <- NA
df$laterality <- relevel(factor(df$laterality, ordered = FALSE), ref = "left")

df$lymph_nodes <- factor(ifelse(df$lymph_nodes_examined_positive > 0, "pos", "neg"), ordered = TRUE, levels = c("neg", "pos"))
df$lymph_nodes[is.na(df$lymph_nodes_examined_positive)] <- NA
df$lymph_nodes <- relevel(factor(df$lymph_nodes, ordered = FALSE), ref = "neg")

df$TMB_count_median <- factor(df$TMB_count_median, ordered = TRUE, levels = c("Lo", "Hi"))
df$TMB_count_median <- relevel(factor(df$TMB_count_median, ordered = FALSE), ref = "Lo")

df$TMB10 <- factor(df$TMB_count_group, ordered = TRUE, levels = c("low (<90%)", "high (>90%)"))
df$TMB10 <- relevel(factor(df$TMB10, ordered = FALSE), ref = "low (<90%)")

df$receptor_subtype <- relevel(factor(df$receptor_subtype, ordered = FALSE), ref = "HR+HER2-")

df$stage <- as.character(df$stage)
df$stage[df$stage == "0"] <- NA
df$stage[df$stage == "1"] <- "I"
df$stage[df$stage == "2"] <- "II"
df$stage[df$stage == "3"] <- "III"
df$stage[df$stage == "4"] <- "IV"
df$stage <- relevel(factor(df$stage, ordered = FALSE), ref = "I")

mb$data <- df
mb$data$histological_subtype[mb$data$histological_subtype %in% c("IDC-TC", "MC", "MPC", "MUC")] <- "other"
mb$data$histological_subtype[is.na(mb$data$histological_subtype)] <- "other"

mb$data$grade_hi <- NA
mb$data$grade_hi[mb$data$grade %in% c(1,2)] <- "low"
mb$data$grade_hi[mb$data$grade %in% c(3,4)] <- "high"

mb$data$receptor_group <- "other"
mb$data$receptor_group[mb$data$receptor_status %in% c("E+", "E+P+")] <- "E+(P+)"
mb$data$receptor_group[mb$data$receptor_status %in% c("H+", "E+H+", "E+P+H+")] <- "H+(E+)(P+)"
mb$data$receptor_group[mb$data$receptor_status %in% c("---")] <- "---"
mb$data$receptor_group[is.na(mb$data$receptor_status)] <- NA

saveRDS(mb, fp(params$data, "mb.rds"))
```


#### Analysis

```{r}
mb <- readRDS(fp(params$data, "mb.rds"))
mb$results <- list()
```

```{r}
mb$data$receptor_subtype |> table(useNA = "always")
mb$data$ER_status |> table(useNA = "always")
mb$data$PR_status |> table(useNA = "always")
mb$data$HER2_status |> table(useNA = "always")
mb$data$histological_subtype |> table(useNA = "always")
mb$data$stage |> table(useNA = "always")
mb$data$grade |> table(useNA = "always")
mb$data$PAM50_subtype |> table(useNA = "always")
mb$data[,c("PAM50_subtype", "HER2_status")] |> table(useNA = "always")
```

```{r}
df <- mb$data
cts <- grep("Cbs_", mb$vars$celltypes, value = T)
x_vars <- setdiff(cts, c("ID", "Cbs_P_value", "Cbs_absolute_score"))
x_vars <- c("ImmuneScore", "StromalScore", x_vars)
for (var in x_vars){
  df[[var]] <- scale(df[[var]])
}
```

Multivariate analysis of all patients
```{r}
# variables
co_vars <-  c("stage", "age", "receptor_subtype") # covariates
y_vars = c("os", "dss", "dfs", "stage_hi", "ER_status", "PR_status", "HER2_status") # survival/clinical

# testing
mb$results$multivar$all <- lapply(setNames(y_vars, y_vars), function(var1){
    lapply(setNames(x_vars, x_vars), function(var2){
      tmp_covars <- setdiff(co_vars, unique(c(var1, var2)))
      tmp_covars <- tmp_covars[get_n_unique(df, tmp_covars) > 1]
        if (any(c(var1) %in% c("ER_status", "PR_status", "HER2_status"))){
            tmp_covars <- setdiff(tmp_covars, "receptor_subtype")
          }
        if (var1 %in% c("stage_hi")){
            tmp_covars <- setdiff(tmp_covars, "stage")
          }
      get_association(df, test_var = var2, ref_var = var1, co_vars = tmp_covars, verbose = FALSE)
    })
})

# pvalue adjustment
padj <- sapply(mb$results$multivar$all, function(var1){
  sapply(var1, function(var2){ var2$res[,"p.value"] })
}) |> padjust()
res_padj <- padj |> rownames2col(col = var) |> tidyr::pivot_longer(-c(var), names_to = "endpoint", values_to = "padj")

# confidence intervals
ep <- c("os", "dss", "dfs")
cox_df <- lapply(setNames(ep, ep), function(e){
  vars <- names(mb$results$multivar$all[[e]])
  tmp <- sapply(setNames(vars, vars), function(var){
      tmp <- broom::tidy(mb$results$multivar$all[[e]][[var]]$fit, conf.int = TRUE, exponentiate = TRUE) |> subset(term == var) |> select(-term)
      unlist(tmp[1, ], use.names = TRUE)
  }) |> t() |> rownames2col(var)
  tmp$endpoint <- e
  tmp
}) |> Reduce(f = rbind)

cox_df <- dplyr::left_join(cox_df, res_padj)
cox_df$signif <- ifelse(cox_df$padj <= 0.05, "signif.", "not signif.") |> factor(ordered = TRUE, levels = c("not signif.", "signif."))
cox_df$var <- factor(cox_df$var, ordered = TRUE, levels = rev(unique(c(cox_df$var))))
cox_df$conf.high[cox_df$conf.high > 100] <- Inf
mb$results$multivar$all_results <- cox_df
```

```{r}
# check proportional hazards assumption
# stratified cox models can be fitted by providing get_association(strat_vars=c("receptor_subtype", ...)), which here only marginally affects the results
zph <- mb$results$multivar$all[c("os", "dss", "dfs")] %S>% function(var1){
  var1 %S>% function(var2){
    survival::cox.zph(var2$fit)$table[1,"p"]
  }
}
zph |> cxheatmap()

# nothing relevant for os, dss; check for dfs
mb$results$multivar$all$dfs$StromalScore$fit |> survival::cox.zph() |> plot()
mb$results$multivar$all$dfs$Cbs_Mast_cells_resting$fit |> survival::cox.zph() |> plot()
```


Multivariate analysis of patient subgroups
```{r}
mc_vars <- c("Cbs_Mast_cells_resting", "Cbs_Mast_cells_activated")
y_vars = c("os", "dss", "dfs", setdiff(x_vars, mc_vars))

# subgroups
mb_subgroup_data <- list(receptor_subtype = split(df, df$receptor_subtype),
                         ER_status = split(df, df$ER_status),
                         HER2_status = split(df, df$HER2_status),
                         PR_status = split(df, df$PR_status),
                         histological_subtype = split(df, df$histological_subtype))
mb_subgroup_data <- unlist(mb_subgroup_data, recursive = FALSE)
mb_subgroup_data <- mb_subgroup_data[sapply(mb_subgroup_data, nrow) > 100]

# testing
mb$results$multivar$subgroups <- mb_subgroup_data %L>% function(df_subgroup){
  subgroup_var <- sub("\\..*", "", .name)
  lapply(setNames(y_vars, y_vars), function(var1){
    lapply(setNames(mc_vars, mc_vars), function(var2){
      tmp_covars <- setdiff(co_vars, unique(c(subgroup_var, var1, var2)))
      tmp_covars <- tmp_covars[get_n_unique(df_subgroup, tmp_covars) > 1]
        if (any(c(var1, subgroup_var) %in% c("ER_status", "PR_status", "HER2_status"))){
            tmp_covars <- setdiff(tmp_covars, "receptor_subtype")
          }
        if (var1 %in% c("stage_hi")){
            tmp_covars <- setdiff(tmp_covars, "stage")
          }
      get_association(df_subgroup, test_var = var2, ref_var = var1, co_vars = tmp_covars, verbose = FALSE)
    })
  })
}

# pvalue adjustment
padj <- lapply(mb$results$multivar$subgroups, function(res){
  padj <- sapply(res, function(var1){
   sapply(var1, function(var2){ if (!any(is.na(var2$res))){ var2$res[,"p.value"] } else NA })
  }) |> padjust()
  padj |> rownames2col(col = xvar) |> tidyr::pivot_longer(-c(xvar), names_to = "var", values_to = "padj")
})
padj <- padj %L>% dplyr::mutate(subgroup = .name)
padj <- Reduce(f = rbind, padj)

# model estimates
df_na <- data.frame(
  estimate = NA,
  std.error = NA,
  statistic = NA,
  p.value = NA,
  conf.low = NA,
  conf.high = NA
)
xvar <- "Cbs_Mast_cells_resting"
cox_df_mc_resting <- mb$results$multivar$subgroups %L>% function(group_res){
  res <- group_res %S>% function(var){
    varname <- .name
    fit <- var[[xvar]]$fit
    res <- df_na
    if (!all(is.na(fit))){
      if (tolower(varname) %in% c("os", "dss", "dfs")){
        res <- broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE) |> dplyr::filter(term == xvar) |> select(-term)
      } else {
        res <- broom::tidy(fit, conf.int = TRUE, exponentiate = FALSE) |> dplyr::filter(term == .name) |> select(-term)
      }
    }
    setNames(as.numeric(res), colnames(res))
  }
  res <- as.data.frame(t(res)) |> rownames2col(var)
  res$subgroup <- .name
  res
}
cox_df_mc_resting <- cox_df_mc_resting|> Reduce(f = rbind)
cox_df_mc_resting <- dplyr::left_join(cox_df_mc_resting, subset(padj, xvar == "Cbs_Mast_cells_resting"))
cox_df_mc_resting$group <- cox_df_mc_resting$subgroup |> sub(pattern="\\..*", replacement = "")
cox_df_mc_resting$level <- cox_df_mc_resting$subgroup |> sub(pattern=".*\\.", replacement = "")
cox_df_mc_resting$level <- factor(cox_df_mc_resting$level, levels=rev(c("HR+HER2-", "HER2+", "TNBC", "ER+", "ER-", "PR+", "PR-", "HER2-", "IDC", "ILC", "other")))
mb$results$multivar$subgroups_results_mc_resting <- cox_df_mc_resting

xvar <- "Cbs_Mast_cells_activated"
cox_df_mc_activated <- mb$results$multivar$subgroups %L>% function(group_res){
  res <- group_res %S>% function(var){
    varname <- .name
    fit <- var[[xvar]]$fit
    res <- df_na
    if (!all(is.na(fit))){
      if (tolower(varname) %in% c("os", "dss", "dfs")){
        res <- broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE) |> dplyr::filter(term == xvar) |> select(-term)
      } else {
        res <- broom::tidy(fit, conf.int = TRUE, exponentiate = FALSE) |> dplyr::filter(term == .name) |> select(-term)
      }
    }
    setNames(as.numeric(res), colnames(res))
  }
  res <- as.data.frame(t(res)) |> rownames2col(var)
  res$subgroup <- .name
  res
}
cox_df_mc_activated <- cox_df_mc_activated |> Reduce(f = rbind)
cox_df_mc_activated <- dplyr::left_join(cox_df_mc_activated, subset(padj, xvar == "Cbs_Mast_cells_activated"))
cox_df_mc_activated$group <- cox_df_mc_activated$subgroup |> sub(pattern="\\..*", replacement = "")
cox_df_mc_activated$level <- cox_df_mc_activated$subgroup |> sub(pattern=".*\\.", replacement = "")
cox_df_mc_activated$level <- factor(cox_df_mc_activated$level, levels=rev(c("HR+HER2-", "HER2+", "TNBC", "ER+", "ER-", "PR+", "PR-", "HER2-", "IDC", "ILC", "other")))
mb$results$multivar$subgroups_results_mc_activated <- cox_df_mc_activated
```

```{r}
zph <- mb$results$multivar$subgroups %L>% function(x){
  res <- x[c("os", "dss", "dfs")] %S>% function(var1){
    var1 %S>% function(var2){
      survival::cox.zph(var2$fit)$table[1,"p"]
    }
  }
  res <- as.data.frame(res) |> rownames2col(var)
  res$group <- .name
  res
} 
Reduce(f = rbind, zph) |> dplyr::arrange(os)

# check IDC
mb$results$multivar$subgroups$histological_subtype.IDC$os$Cbs_Mast_cells_resting$fit |> survival::cox.zph() |> plot()
```

DEG analysis for subgroups
```{r}
if (!file.exists(fp(params$data, "genesets.rds"))){
  gs <- getMSigDB(c("H", "C2|CP:KEGG_LEGACY", "C5|GO:BP|GO:CC|GO:MF"))
} else {
  gs <- readRDS(fp(params$data, "genesets.rds"))
}
cbs_sig <- read.delim("../lib/cibersort/LM22.txt")
cbs_sig$gene <- cbs_sig$Gene.symbol
rownames(cbs_sig) <- cbs_sig$Gene.symbol
```

Prepare data for DGEA
```{r}
dfde <- mb$data
dfde$receptor_status <- dfde$receptor_status |> make.names()
dfde$receptor_subtype <- dfde$receptor_subtype |> make.names()
dfde$stage <- dfde$stage |> make.names()
dfde$HER2_status_orig <- dfde$HER2_status
dfde$HER2_status <- dfde$HER2_status_orig |> gsub(pattern = "\\+", replacement = "pos") |> gsub(pattern = "-", replacement = "neg")
dfde$HER2_status[is.na(dfde$HER2_status)] <- "other"
dfde$histological_subtype <- dfde$histological_subtype |> make.names()
rownames(dfde) <- dfde$ID
```

DGEA and GSEA
```{r}
dfde$HR_status <- (((dfde$ER_status == "ER+") | (dfde$PR_status == "PR+")) & (dfde$HER2_status == "HER2neg")) |> ifelse("HRpos", "HRneg")
table(dfde$HR_status)
table(dfde[,c("HR_status", "ER_status")])
table(dfde[,c("HR_status", "PR_status")])
table(dfde[,c("HR_status", "HER2_status")])
dfsub <- subset(dfde, HR_status == "HRpos")
table(dfsub$receptor_subtype)
dfsub <- dfsub |> select(ID, Cbs_Mast_cells_resting, Cbs_Mast_cells_activated, stage, age, grade, histological_subtype) |> tidyr::drop_na()
dfsub$Mast_cells_resting_hilo <- dfsub$Cbs_Mast_cells_resting |> stratify(n = 2)
dfsub$Mast_cells_activated_hilo <- dfsub$Cbs_Mast_cells_activated  |> stratify(n = 2)

# resting
res <- runLIMMA(mb$expr[,dfsub$ID], design = dfsub,
                formula = ~ Mast_cells_resting_hilo + stage + age + grade + histological_subtype,
                contrasts = list("Mast_cells" = c("Mast_cells_resting_hilo", "Hi", "Lo")), do_log = FALSE)
mb$results$dgea$mc_resting <- res$results$Mast_cells
mb$results$dgea$mc_resting$Cbs_score <- cbs_sig[mb$results$dgea$mc_resting$id,"Mast.cells.resting"]
mb$results$dgea$mc_resting$Cbs_score[is.na(mb$results$dgea$mc_resting$Cbs_score)] <- 0
mb$results$dgea$mc_resting$name <- mb$results$dgea$mc_resting$id
mb$results$gsea$mc_resting <- mb$results$dgea$mc_resting |> runGSEA(rank_by = "log2FC", genesets = gs)

# activated
res <- runLIMMA(mb$expr[,dfsub$ID], design = dfsub,
                formula = ~ Mast_cells_activated_hilo + stage + age + grade + histological_subtype,
                contrasts = list("Mast_cells" = c("Mast_cells_activated_hilo", "Hi", "Lo")), do_log = FALSE)
mb$results$dgea$mc_activated <- res$results$Mast_cells
mb$results$dgea$mc_activated$Cbs_score <- cbs_sig[mb$results$dgea$mc_activated$id,"Mast.cells.activated"]
mb$results$dgea$mc_activated$Cbs_score[is.na(mb$results$dgea$mc_activated$Cbs_score)] <- 0
mb$results$dgea$mc_activated$name <- mb$results$dgea$mc_activated$id
mb$results$gsea$mc_activated <- mb$results$dgea$mc_activated |> runGSEA(rank_by = "log2FC", genesets = gs)
```

```{r}
saveRDS(mb$results, fp(params$data, "mb_results.rds"))
```
